Cognate beginnings to bilingual lexical acquisition

### 2023-07-03 manuscript/manuscript.qmd
Add comments by DAV

---
title: "Cognate beginnings to bilingual lexical acquisition"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        city: Barcelona, 08005
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    email: avila.varela.daniela@gmail.com
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        city: Lisboa, 1600-214
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    email: jignaciocastillejo@gmail.com
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        city: Madrid, 28049
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    email: nuria.sebastian@upf.edu
    affiliations: 
      - ref: cbc
keywords:
  - cognate
  - word acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source
lang: en

format:
  arxiv-pdf:
    keep-tex: false  
    linenumbers: true
    doublespacing: true
    runninghead: "Cognate beginnings to bilingual lexical acquistion"
    include-in-header: 
      text: |
        \usepackage{tipa}
        \usepackage{booktabs}
abstract: |
    Bilingual infants show equivalent developmental trajectories of lexical acquisition compared to their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalise on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
fig-dpi: 1000
---

@@

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of language distance on vocabulary size.

@@

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

@@

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated^[We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not make strong assumptions about whether a learning instance is a discrete or a continuous unit of accumulation of information. We rather consider that a learning instance of a word is so to the degree of the strength of activation of its phonological representation. This activation may be the result of the infant being exposed to the word-form, or the result of activation spreading through phonological or semantic links across lexicl representation, as in the case of parallel activation.]. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive additional activation through parallel activation more often than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], that  are phonologically very similar. When the child is exposed to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. Therefore, this exposure will count as a learning instance for both co-activated forms. However, for the case of the non-cognate translation equivalent /\textipa{"gos}/, the low phonological similarity will result in a weak activation of /\textipa{"pe.ro}/ resulting in such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. We adopted a Bayesian explanatory item response theory approach [IRT, see @kachergis2022standard, for a similar approach] to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition, including participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency. We predicted an interaction between cognate status and word exposure rate  in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure words cognate words.

@@

```{r tbl-items}
#| label: tbl-items
#| tbl-cap: "Summary of the items included in the final analyses."
#| eval: true
#| echo: false
#| warning: false
items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(total = rowSums(across(a:d)),
           items = map(items, function(x) tail(x, 2))) |>  
    arrange(desc(total)) |> 
    relocate(a:d, total, items) |> 
    mutate(group = "Semantic category") |> 
    select(-total) |> 
    gt(groupname_col = "group",
       rowname_col = "semantic_category") |> 
    fmt_percent(matches("_prop"), decimals = 1) |> 
    cols_label(a = "List A",
               b = "List B",
               c = "List C",
               d = "List D",
               items = "Examples") |> 
    summary_rows(groups = "Semantic category",
                 columns = a:d,
                 fns = list("Total" = ~sum(., na.rm = TRUE))
    ) |> 
    tab_style(cell_text(size = "small"),
              list(cells_body(),
                   cells_summary(),
                   cells_stub(),
                   cells_column_labels(),
                   cells_column_spanners(),
                   cells_stub_summary(),
                   cells_row_groups())) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_options(data_row.padding = px(2),
                row_group.padding = px(1.2)) |> 
    tab_style(cell_text(size = "x-small", align = "left"),
              cells_body(items))

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants) for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4-month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
ftn <- "For participants exposed to Catalan and Spanish exclusively this proportion is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    sub_missing(everything(), everything(), missing_text = "-") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:7) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan))) |>
    tab_footnote(ftn, location = cells_column_labels(doe_catalan))
```

@@

We collected responses to `r format(nrow(n_items), big.mark = ",")` items. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to words `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest: *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (a language exposure-weighted lexical frequency), and *Cognateness* (defined as the phonological similarity between translation equivalentsFor each translation equivalent). We added these variables as main effects, together with the two-way and three-way interactions between *Age*, *Exposure*, and *Cognateness*.

>>>>

---
title: "Cognate beginnings to bilingual lexical acquisition"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        city: Barcelona, 08005
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    email: avila.varela.daniela@gmail.com
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Linguistics Center (CLUL), School of Arts and Humanities
        city: Lisbon, 1600-214
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    email: jignaciocastillejo@gmail.com
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        city: Madrid, 28049
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    email: nuria.sebastian@upf.edu
    affiliations: 
      - ref: cbc
keywords:
  - cognate
  - word acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source
lang: en

format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: true
    doublespacing: false
    runninghead: "Cognate beginnings to bilingual lexical acquistion"
    include-in-header: 
      text: |
        \usepackage{tipa}
        \usepackage{booktabs}
abstract: |
    Bilingual infants show equivalent developmental trajectories of lexical acquisition compared to their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalise on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. GGC was supported by the This research was supported by a FPI 2020 research contract (PRE2019-088165) in a project funded by the Spanish Ministry of Science, Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported by the European Union's Horizon 2023 research and innovation program under Marie Skłodowska–Curie Grant (765556) and a postdoctoral fellowship from the Foundation for Science and Technology of Portugal (UIDB/00214/2020). NSG was supported by Academia 2019 award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
fig-dpi: 1000
---

@@

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degre of cognateness) on vocabulary size.

@@

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

@@

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated^[We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not make strong assumptions about whether a learning instance is a discrete or a continuous unit of accumulation of information. We rather consider that a learning instance of a word is so to the degree of the strength of activation of its phonological representation. This activation may result from the infant being exposed to the word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation.]. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language (i.e., the language to which the infants are exposed to more often) would receive additional activation through parallel activation more frequently than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], that  are phonologically very similar. When the child is exposed to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos-"pe.ro}/ would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq], a dual language online vocabulary checklist designed for the present study. We adopted a Bayesian explanatory item response theory approach [IRT, see @kachergis2022standard, for a similar approach] to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition, including participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency. We predicted an interaction between cognate status and word exposure rate  in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure cognate words.

@@

```{r tbl-items}
#| label: tbl-items
#| tbl-cap: "Summary of the items included in the final analyses."
#| eval: true
#| echo: false
#| warning: false
col_names <- c("Semantic category", 
               "List A",
               "List B",
               "List C",
               "List D",
               "Examples")

tbl_data <- items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(total = rowSums(across(a:d)),
           items = map(items, function(x) tail(x, 2))) |>  
    arrange(desc(total)) |> 
    relocate(semantic_category, a:d, total, items) |> 
    select(-total)

sum_rows <- function(x) {
    if (is.numeric(x)) sum(x) 
    else if(is.list(x)) ""
    else "N"
}

sums_row <- as.data.frame(lapply(tbl_data, sum_rows),
                          col.names = colnames(tbl_data))
tbl_data <- rbind(tbl_data, sums_row)

knitr::kable(tbl_data,
             format = "latex", 
             digits = 1, 
             booktabs = TRUE,
             col.names = col_names,
             align = "lccccl")  |> 
    kableExtra::row_spec(nrow(tbl_data)-1, 
                         hline_after = TRUE)
```

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "For participants exposed to Catalan and Spanish exclusively, the proportion of Catalan exposure shown in the table is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%. Participants with 100\\%, 0\\%, and 50\\% Catalan exposure would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively."
ftn <- ""

col_names <- c("Catalan exposure",
               "[10-14]",
               "(14, 18]",
               "(18, 22]", 
               "(22-26]", 
               "(26-30]", 
               "(30-34]")

options(knitr.kable.NA = 0)

tbl_data <- participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, 
                values_from = n,
                names_repair = janitor::make_clean_names) |> 
    mutate(across(where(is.numeric), \(x) ifelse(is.na(x), 0, x)))

sum_rows <- function(x) if (is.numeric(x)) sum(x) else "N"
sums_row <- as.data.frame(lapply(tbl_data, sum_rows),
                          col.names = colnames(tbl_data))
tbl_data <- rbind(tbl_data, sums_row)

knitr::kable(tbl_data,
             format = "latex", 
             digits = 1, 
             booktabs = TRUE,
             col.names = col_names,
             align = "lcccccc")  |> 
    kableExtra::add_header_above(c(" " = 1,
                                   "Age (months)" = ncol(tbl_data)-1)) |>
    kableExtra::row_spec(nrow(tbl_data)-1, 
                         hline_after = TRUE)
```

@@

### Data processing

We collected `r format(n_obs, big.mark = ",")` responses to `r format(nrow(n_items), big.mark = ",")` items. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

### Modelling approach

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest: *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (a language exposure-weighted lexical frequency), and *Cognateness* (defined as the phonological similarity between translation equivalents). We added these variables as main effects, together with the two-way and three-way interactions between *Age*, *Exposure*, and *Cognateness*. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [@barr2013random]. We specified a weakly informative prior around the parameters of the model. @eq-linear shows a detailed description of the model.
